---
title: 'The Effects of Heavy Drinking, Smoking, Anxiety and Depression on Stroke Odds:
  A Cross-Sectional Study'
author:
- name: Robert Wardrup
- name: Elizabeth Murphy
- name: Natalie Requinez
- name: Joshua Averitt
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: ../Data/Misc/svm-latex-ms.tex
  word_document: default
  html_document:
    df_print: paged
biblio-style: apa
fontfamily: mathpazo
fontsize: 11pt
geometry: margin=1in
keywords: Stroke, Alcohol, Smoking, Anxiety, Depression
bibliography: ../Data/Misc/bios.bib
thanks: ''
abstract: Though numerous studies have investigated the relationships between lifestyle
  risk factors and stroke outcome, few have done so while controlling as age and race/ethnicity
  range. Utilizing quasi-binomial logistic regression, we show that the odds of having
  had a stroke increase with being a daily smoker, days of anxiety, and having a depression
  diagnosis.
---

```{r init, echo = F, message = F, warning = F}
library(tidyverse)
library(knitr)
#library(rmarkdown)
#library(pander)
#library(DT)

results <- readRDS("/storage/Research/bios5300/Data/results.rds")
colorPalette <- results$colorPalette

opts_chunk$set(echo = F,
               message = F,
               warning = F,
               error = F)
knitr::opts_chunk$set(fig.width = 12,
                      fig.height = 8)
```

# Introduction
Natalie's text here.

``` {r figWithCaption, fig.cap="2016 United States Stroke Prevalence", echo=F}
results$map.stroke.prevalence
```

# Methods
## Josh's Text
## Survey Weighting
The BRFSS survey data were weighted using the raking method, which is a two-part
methodology to help insure unbiased results by accounting for noncoverage and 
nonresponse bias and forcing the total number of cases to equal the population 
estimate of each state in the United States [@CentersforDiseaseControlandPrevention2007a].
Raking works by repeatedly adjusting weight across a set of selected variables until
the weights converge and the survey population totals are equal to the census population
totals for each selected variable [@Fricker1993].

## Dependent and Independent Variables
The dependent variable, stroke outcome, was recoded to a binary variable for use 
in a logistic regression model. Depression diagnosis and heavy drinker were also 
recoded into a binary variable. For each variable used in the analysis, 
*don't know / unsure* and *refused* responses were dropped. Table 1 lists the 
variables included in the model.

Variable Name         | Variable Type
-------------         | -------------
Stroke Diagnosis      | Dichotomous (DV)
Heavy Alcohol Drinker | Dichotomous
Depression Diagnosis  | Dichotomous
Sex                   | Dichotomous
Days Anxious          | Continuous
Daily Sleep Hours     | Continuous
Smoker Status         | Categorical
Race/Ethnicity        | Categorical
Age Group             | Categorical

Table: Dependent and Independent Variables Used in Logistic Regression Model

A subject was designated as a heavy drinker if they were either an adult male
who reported consuming more than 14 drinks per week or an adult female who 
reported consuming more than 7 drinks per week. This value was calculated after
asking the subjects, "during the past 30 days, on the days when you drank, about
how many drinks did you drink on average?" and "during the past 30 days, how many
days per week or month did you have at least one drink of any alcoholic beverage
such as beer, wine, a malt beverage or liquor?"
[@CentersforDiseaseControlandPrevention2007].

## Logistic Regression Model
The first model used in calculating the logistic regression utilized the 
standard binomial function to predict the log odds of binary outcome $k$. The 
binomial formula, is given as
$$P(k)={n \choose k}p^{k}(1-p)^{n-k},$$
where $k$ represents a stroke outcome, $n$ represents the sample size, and $p$ 
is the probability that a stroke will occur. [**Explain overdispersion**] 
Before being checked for overdispersion by calculating $\phi$, given by the 
formula
$$\phi = \frac{1}{(n-p-1)}\sum_{i=1}^{n} (y_{i} -\hat{y} _{i})^{2} 
/ \hat{y} _{i}.$$
A threshold of $\phi > 1$, was used to determine if the data were overdispersed, 
possibly leading to unstable estimates. To account for the overdispersion, 
$\phi$ was included as a model parameter, giving the formula
$$P(k)={n \choose k}p(p+k\phi)^{k-1}(1-p-k\phi)^{n-k}.$$
[**Explain the quasi-binomial family more**]

# Results

The weighted number of strokes in the United States in 2016 was 8,020,080, 
representing approximately 2.5% of the total 2016 United States population. 
Out of these, 3.20% were female and 3.13% were male. 

Diagnosis   | Male    | Female
---------   | ----    | ------
No Stroke   | 96.87%  | 96.80%
Stroke      | 3.13%   | 3.20%

Table: Proportion of Stroke Outcomes in Males and Females

A chi-squared test of independence revealed that there there is not statistically
significant evidence that stroke outcome and sex is independent from one another
at an alpha level of 0.05, $\chi^2$ (3, $N$=486,303) = 1.73, $p$=0.48.

The proportion of subjects who have suffered from a stroke by smoker status is
shown in Table 3, and visually in Figure 2. The percentage of subjects who have
had a stroke is highest for the subjects who smoke daily, followed by subjects who
are former smokers, followed by subjects who smoke some days. A possible
explanation for the lower percentage of stroke in subjects who smoke some days than
subjects who are former smokers is that stroke victims may simply have ceased
smoking after having a stroke. 

Smoker Status     | No Stroke | Stroke
------------------|-----------|-------
Never Smoked      | 97.8%     | 2.2%
Former Smoker     | 95.4%     | 4.6%
Smokes Some Days  | 96.2%     | 3.8%
Smokes Daily      | 95.0%     | 5.0%

Table: Proportion of Stroke Outcomes and Smoker Status

The proportions shown in Table 3 are depicted as a mosaic plot in Figure 2. The
mosaic plot is a graphical depiction the proportions within each table cell,
shaded by the difference from the expected observation. Blue depicts a higher than
expected number of observations for that cell, and red depicts a lower than
expected number of observations for that cell. As shown in Figure 2, there were 
a greater than expected number of non-smokers that have never had a stroke, and 
a greater than expected number of smokers and former smokers who have had a 
stroke. 

``` {r figWithCaption3, fig.cap="Proportion of Stroke Outcomes and Smoker Status", echo=F}
library(vcd)
mosaic(results$stroke.smoker, shade = T,
       main = "Proportions of Stroke Diagnoses by Smoker Status",
       spacing = spacing_equal(1.5),
       keep_aspect_ratio = T,
       labeling_args = list(set_varnames = c(diag.stroke="Stroke Outcome", smoker.status="Smoker Status")))
```

A chi-squared test of independence revealed that there is statistically
significant evidence that stroke outcome and smoking status were not independent
from one another, $\chi^2$ (3, $N$=486,303) = 2352.5, $p <$ 0.001.

The mean hours of sleep for those who had a stroke was 6.99 hours and 6.98 for 
those who did not have a stroke. 

Factor                | Outcome     | Mean
--------              | -------     | ----
Days Anxious          | No Stroke   | 4.79
                      || Stroke     | 8.05
Nightly Sleep Hours   | No Stroke   | 6.98
                      || Stroke     | 6.99
Number Drinks Weekly  | No Stroke   | 79.07
                      || Stroke     | 40.31

Table: Mean Values of Selected Risk Factors
                      

## Logistic Regression Model
The binomial logistic regression model resulted in a $\phi$ value of 2.52, 
indicating that there is overdispersion in the estimates and the quasi-binomial 
function should be used instead.

The results of the quasi-binomial logistic regression (Table 4) indicate that,
controlling for age, race/ethnicity, and sex, Several of the tested risk factors
were significantly correlated with stroke outcome. Being a daily smoker,
Monthly days with anxiety, and diagnosed depression were all positively 
associated with greater odds of stroke.

Daily cigarette smokers had 1.68 times the odds of having had a stroke than 
those who have never smoked cigarettes. For each additional day heavily affected 
by anxiety,the odds of having had a stroke were 1.03 times higher, and for those 
who have been diagnosed with depression the odds of stroke were 2.12 times 
higher. Of the control variables, age of 25 years old and above were the only 
factors significantly associated with stroke outcome, with an odds ratio
that increased with age. 

Variable                                                  | OR (95% CI)             | P-Value
------------                                              | --------------------    | -------
**Risk Factors**                                          ||
&nbsp;&nbsp;&nbsp;*Drinking*                              ||
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Non-Heavy Drinker           | 1 (Baseline Factor)     |
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Heavy Drinker               | 0.989 (0.41 - 2.38)     | 0.98
&nbsp;&nbsp;&nbsp;*Smoking*                               ||
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Non-Smoker                  | 1 (Baseline Factor)     |
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Former Smoker               | 1.24 (0.91 - 1.70)      | 0.18
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Smoker (Some days)          | 1.65 (0.95 - 2.88)      | 0.08
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Smoker (Daily)              | 1.68 (1.12 - 2.53)      | 0.01
&nbsp;&nbsp;&nbsp;*Mental Health*                         ||
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Monthly Days Anxious        | 1.03 (1.01 - 1.04)      | <0.01
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;No Depression Diagnosis     | 1 (Baseline Factor)     |
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Depression Diagnosis        | 2.12 (1.48 - 3.04)      | <0.01
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Daily Sleep Hours           | 0.91 (0.82 - 1.00)      | 0.06
&nbsp;&nbsp;&nbsp;*Sex*                                   ||
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Male                        | 1 (Baseline Factor)     |
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;Female                      | 0.77 (0.58 - 1.02)      | 0.07
**Race/Ethnicity**                                        ||
&nbsp;&nbsp;&nbsp;Other Race / Non-Hispanic               | 1 (Baseline Factor)     |
&nbsp;&nbsp;&nbsp;Hispanic                                | 1.42  (0.43 - 4.70)     | 0.57
&nbsp;&nbsp;&nbsp;Black Only / Non-Hispanic               | 2.41  (0.88 - 6.62)     | 0.09
&nbsp;&nbsp;&nbsp;White Only / Non-Hispanic               | 1.87  (0.73 - 4.83)     | 0.20
&nbsp;&nbsp;&nbsp;Multiracial / Non-Hispanic              | 2.92  (0.90 - 9.50)     | 0.08
**Age Group**                                             ||
&nbsp;&nbsp;&nbsp;Age 18 to 24                            | 1 (Baseline Factor)     |
&nbsp;&nbsp;&nbsp;Age 25 to 34                            | 3.92 (0.80 - 19.66)     | 0.10
&nbsp;&nbsp;&nbsp;Age 35 to 44                            | 9.33 (2.00 - 43.51)     | <0.01
&nbsp;&nbsp;&nbsp;Age 45 to 54                            | 12.35 (2.79 - 54.71)    | <0.01
&nbsp;&nbsp;&nbsp;Age 55 to 64                            | 20.93 (4.80 - 91.34)    | <0.01
&nbsp;&nbsp;&nbsp;Age 65 or Older                         | 44.87 (10.36 - 194.43)  | <0.01

Table: Logistic Regression Model Estimating Effects of Risk Factors & 
Demographic Variables on Stroke Outcome Odds.

The effect of each significant risk factor, stratified by depression diagnosis, 
on the probability of having had a stroke is shown in Figure 2. The 
probability of a person having a stroke increases with the number of days a
person is significantly affected by anxiety. This probability is multiplied if
a person has been diagnosed with depression. For non-smokers with no depression
diagnosis who did not signifantly suffer from anxiety, the probability of having
had a stroke approaches zero. If the same subject was diagnosed with depression,
the probability of having had a stroke nears 5%. 

The probability of having had a stroke is multiplied in daily smokers, beginning
at approximately 5% for daily smokers who have not been diagnosed with 
depression nor have suffered from significant anxiety. With a depression
diagnosis, the probability originates at approximately 7% for subjects with zero
days of anxiety and increases to 15% for subjects who experienced thirty days of
signiticant anxiety. This relationship is similar for subjects who currently
smoke some days.

``` {r figWithCaption2, fig.cap="Effects of Anxious Days on Stroke Outcome, Stratified by Depression Diagnosis and Smoker Status", echo=F}
plot(results$effects.anxiety.depression.smoking, 
     multiline=T, 
     main = "Effect of Anxious Days on Stroke Outcome, Stratified by Depression Diagnosis and Smoker Status",
     xlab = "Anxious Days",
     ylab = "Probability of Having Had a Stroke",
     ci.style="band")
```

# Conclusion
Our study assessed the association between heavy drinking, anxiety, depression on stroke odds. We used the 2016 Behavioral Risk Factor Surveillance System dataset from the CDC. There was a significant difference between being a daily smoking and odds of a stroke as the p-value was .01. Also, there was a significant difference for anxiety, and depression for the odds of a stroke, p-value was less than .01 for both variables. We found know no significant different between race or the amount of sleep hours someone receives., and whether or not an individual is a heavy drinker. 

Our study has a specific limitation due to its study design. The study was a cross-sectional one, causality and risk cannot be inferred from our data. Each variable tested affected one another, so heavy drinking influenced the variables anxiety and depression. And anxiety can affect heavy drinking and depression. The influence of variables on each other demonstrates that causality cannot be inferred in a cross-sectional study. 

In conclusion, variables depression, anxiety, and being a daily smoker all were significantly different for stroke odds. This conclusion means that an individual who is either a daily smoker, diagnosed with depression or anxiety all have an increased change of having a stroke. The 2016 BRFSS dataset provided a large population data set to help monitor variables that affect stroke odds. Future studies can look at the association can look further at the effect of mental health diagnosis on stroke odds, as well as how smoking can increase an individual’s odds of a stroke. 
