---
title: 'The Effects of Heavy Drinking, Smoking, Anxiety and Depression on Stroke 
  Odds: A Cross-Sectional Study'
author:
- name: Robert Wardrup
- name: Elizabeth Murphy
- name: Natalie Requinez
- name: Joshua Averitt
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: ../Data/Misc/svm-latex-ms.tex
  html_document:
    df_print: paged
biblio-style: apa
fontfamily: mathpazo
fontsize: 11pt
geometry: margin=1in
keywords: Stroke, Alcohol, Smoking, Anxiety, Depression
bibliography: ../Data/Misc/bios.bib
thanks: ''
abstract: Although numerous studies have investigated the relationshiops between 
  lifestyle risk factors and stroke outcome, few have done so while controlling
  for demographic variables such as age, race/ethnicity, education and income 
  range. Utilizing quasi-binomial logistic regression, we show that females and
  those in higher income brackets have a lower odds of having had a stroke 
  while the odds of having had a stroke increase with being a daily smoker, 
  days of anxiety, and having a depression diagnosis.
---

```{r init, echo = F, message = F, warning = F}
library(tidyverse)
library(knitr)
#library(rmarkdown)
#library(pander)
#library(DT)

results <- readRDS("../Data/results.rds")
colorPalette <- results$colorPalette

opts_chunk$set(echo = F,
               message = F,
               warning = F,
               error = F)
knitr::opts_chunk$set(fig.width = 8,
                      fig.height = 6)
```

# Introduction
Natalie's text here.

Something, something geographic distribution of stroke is clustered in a portion 
of the Southeastern region of the United States. 
in Figure 1.

``` {r figWithCaption, fig.cap="2016 United States Stroke Prevalence", echo=F}
results$map.stroke.prevalence
```

# Methods
## Josh's Text
## Survey Weighting
The BRFSS survey data were weighted using the raking method, which is a two-part
methodology to help insure unbiased results by accounting for noncoverage and 
nonresponse bias and forcing the total number of cases to equal the population 
estimate of each state in the United States [@CentersforDiseaseControlandPrevention2007].

## Dependent and Independent Variables
The dependent variable, stroke outcome, was recoded to a binary variable for use 
in a logistic regression model. Depression diagnosis and heavy drinker were also 
recoded into a binary variable. For each variable used in the analysis, 
*don't know / unsure* and *refused* responses were dropped. 
Variables included in the model were:

Variable Name         | Variable Type
-------------         | -------------
Stroke Diagnosis      | Dichotomous (DV)
Heavy Alcohol Drinker | Dichotomous
Depression Diagnosis  | Dichotomous
Sex                   | Dichotomous
Days Anxious          | Continuous
Daily Sleep Hours     | Continuous
Smoker Status         | Categorical
Education Level       | Categorical
Income Level          | Categorical
Race/Ethnicity        | Categorical
Age Group             | Categorical

Although there were clearly outliers in the daily sleep hours variable, they 
were not removed from the data. The maximum value for daily sleep hours was 
20 and the minimum was 0. The mean, including outliers was 7.05 hours and 
the mean with the outliers removed was 7.01 hours. The mean of the outliers 
themselves was 9.78 hours. 

## Logistic Regression Model
The first model used in calculating the logistic regression utilized the 
standard binomial function to predict the log odds of binary outcome $k$. The 
binomial formula, is given as
$$P(k)={n \choose k}p^{k}(1-p)^{n-k},$$
where $k$ represents a stroke outcome, $n$ represents the sample size, and $p$ 
is the probability that a stroke will occur. [**Explain overdispersion**] 
Before being checked for overdispersion by calculating $\phi$, given by the 
formula
$$\phi = \frac{1}{(n-p-1)}\sum_{i=1}^{n} (y_{i} -\hat{y} _{i})^{2} 
/ \hat{y} _{i}.$$
A threshold of $\phi > 1$, was used to determine if the data were overdispersed, 
possibly leading to unstable estimates. To account for the overdispersion, 
$\phi$ was included as a model parameter, giving the formula
$$P(X=k)={n \choose k}p(p+k\phi)^{k-1}(1-p-k\phi)^{n-k}.$$
[**Explain the quasi-binomial family more**]

# Results

The weighted number of strokes in the United States in 2016 was 8,020,080, 
representing approximately 2.5% of the total 2016 United States population. 
Out of these, 3.20% were female and 3.13% were male. 

Diagnosis   | Male    | Female
---------   | ----    | ------
No Stroke   | 96.87%  | 96.80%
Stroke      | 3.13%   | 3.20%

Table: Proportion of Stroke Outcomes in Males and Females

A chi-squared test of proportions revealed that the difference between number of 
strokes for males and females was not statistically significant 
($\chi^2 = 1.73; p = 0.48$).  

The mean hours of sleep for those who had a stroke was 6.99 hours and 6.98 for 
those who did not have a stroke. 

Factor                | Outcome     | Mean
--------              | -------     | ----
Days Anxious          | No Stroke   | 4.79
                      || Stroke     | 8.05
Nightly Sleep Hours   | No Stroke   | 6.98
                      || Stroke     | 6.98
Number Drinks Weekly  | No Stroke   | 79.07
                      || Stroke     | 40.31

Table: Mean Values of Selected Risk Factors
                      

## Logistic Regression Model
###Quasi-Binomial Model Results
The binomial logistic regression model resulted in a $\phi$ value of 2.52, 
indicating that there is overdispersion in the estimates and the quasi-binomial 
function should be used instead.

The results of the quasi-binomial logistic regression (Table 4) indicate that 
being female is a protective factor (OR = 0.71, 95% CI = [0.51 - 0.98]), as well 
as having an annual household income between \$35,000 and \$75,000 
(OR = 0.48, 95% CI = [0.25 - 0.92]) or over $75,000 
(OR = 0.41, 95% CI = [0.20 - 0.92]). 

Monthly days with anxiety, having a depression diagnosis, and age were all 
positively associated with greater odds of stroke.For each additional day 
heavily affected by anxiety,the odds of having had a stroke were 1.02 times 
higher, and for those who have been diagnosed with depression the odds of stroke 
were 2.05 times higher. Age had the strongest relationship to the odds of having 
had a stroke, with those aged over 65 years old having 95 times the odds of 
having had a stroke diagnosis than those who were between the ages of 18 and 24 
years old.

Variable                                      | OR (95% CI)
------------                                  | ------------------
**Risk Factors**                              ||
&nbsp;&nbsp;&nbsp;Non-Heavy Drinker           | 1
&nbsp;&nbsp;&nbsp;Heavy Drinker               | 1.12 (0.48 - 2.62)
&nbsp;&nbsp;&nbsp;Non-Smoker                  | 1
&nbsp;&nbsp;&nbsp;Former Smoker               | 1.06 (0.75 - 1.49)
&nbsp;&nbsp;&nbsp;Smoker (Some days)          | 1.10 (0.60 - 2.02)
&nbsp;&nbsp;&nbsp;Smoker (Daily)              | 1.10 (0.69 - 1.76)
&nbsp;&nbsp;&nbsp;Monthly Days Anxious        | 1.02 (1.01 - 1.04)**
&nbsp;&nbsp;&nbsp;No Depression Diagnosis     | 1
&nbsp;&nbsp;&nbsp;Depression Diagnosis        | 2.05 (1.37 - 3.06)**
&nbsp;&nbsp;&nbsp;Daily Sleep Hours           | 0.90 (0.81 - 1.01)
&nbsp;&nbsp;&nbsp;Male                        | 1
&nbsp;&nbsp;&nbsp;Female                      | 0.71 (0.51 - 0.98)**
**Education Level**                           |
&nbsp;&nbsp;&nbsp;None/Only Kindergarten      | 1
&nbsp;&nbsp;&nbsp;Grades 1 through 8          | 2.06  (0.28 - 15.32)
&nbsp;&nbsp;&nbsp;Grades 9 through 11         | 2.56  (0.34 - 19.33)
&nbsp;&nbsp;&nbsp;Grade 12 or GED             | 1.29  (0.17 - 9.64)
&nbsp;&nbsp;&nbsp;College: 1 to 3 years       | 2.10  (0.29 - 15.34)
&nbsp;&nbsp;&nbsp;College - 4 or more years   | 1.29  (0.17 - 9.64)
**Income**                                    |
&nbsp;&nbsp;&nbsp;<$15,000                    | 1.50  (0.80 - 2.81)
&nbsp;&nbsp;&nbsp;\$15,000 - $20,000          | 1.61  (0.90 - 2.89)
&nbsp;&nbsp;&nbsp;\$20,000 - $25,000          | 0.97  (0.53 - 1.76)
&nbsp;&nbsp;&nbsp;\$25,000 - $35,000          | 0.91  (0.48 - 1.71)
&nbsp;&nbsp;&nbsp;\$35,000 - $50,000          | 0.55  (0.30 - 1.02)
&nbsp;&nbsp;&nbsp;\$50,000 - $75,000          | 0.48  (0.25 - 0.92)**
&nbsp;&nbsp;&nbsp;>$75,000                    | 0.41  (0.20 - 0.92)**
**Race/Ethnicity**                            |
&nbsp;&nbsp;&nbsp;Other Race / Non-Hispanic   | 1
&nbsp;&nbsp;&nbsp;Hispanic                    | 0.88  (0.26 - 3.03)
&nbsp;&nbsp;&nbsp;Black Only / Non-Hispanic   | 1.87  (0.66 - 5.30)
&nbsp;&nbsp;&nbsp;White Only / Non-Hispanic   | 1.75  (0.66 - 4.65)
&nbsp;&nbsp;&nbsp;Multiracial / Non-Hispanic  | 2.69  (0.77 - 9.40)
**Age Group**                                 |
&nbsp;&nbsp;&nbsp;Age 18 to 24                | 1
&nbsp;&nbsp;&nbsp;Age 25 to 34                | 10.46 (1.27 - 86.00)**
&nbsp;&nbsp;&nbsp;Age 35 to 44                | 26.31 (3.40 - 203.40)**
&nbsp;&nbsp;&nbsp;Age 45 to 54                | 28.90 (3.81 - 219.13)**
&nbsp;&nbsp;&nbsp;Age 55 to 64                | 51.08 (6.87 - 379.54)**
&nbsp;&nbsp;&nbsp;Age 65 or Older             | 95.22 (12.74 - 711.66)**

Table: Logistic Regression Model Estimating Effects of Risk Factors & 
Demographic Variables on Stroke Outcome Odds.

The effect of each variable, stratified by depression diagnosis, on the 
probability of having had a stroke is represented in Figure 2. 

# Conclusion
Beth's part here